Assessing landscape N removal in coastal New England catchments using the

### 2023-10-04 manuscript/manuscript.Rmd
Add Rachel as co-author, fix fig 4 (A to B), add note about terra package, edits to acknowledgements,

---
title: "Assessing landscape N removal in coastal New England catchments using the
  N-Sink approach with a new R package, nsink"
author:
- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C. L.^*3*^, Gold, A. J. ^*1*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: "Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water
  quality. Land management decisions made within estuarine watersheds have a direct
  impact on downstream N delivery. Natural features frequently found within the 
  headwaters of watersheds can act as landscape sinks for N, such as wetlands, streams 
  and ponds that transform dissolved N into gaseous N, effectively removing it from 
  the aquatic system. Identifying and evaluating these landscape sinks and their spatial
  relationship to N sources can help managers understand the effects of alternative 
  decisions on downstream resources. N-Sink is an approach that uses widely available 
  GIS data to identify landscape sinks within
  HUC-12 (or larger) catchments, estimate their N removal potential and map the effect
  of those sinks on N movement through the catchment. A series of static maps are
  produced to visualize N removal efficiency, transport and delivery, the latter in
  the form of an index. To facilitate data acquisition, processing and visualization,
  an R package `nsink` was developed and applied to produce static maps for all HUC-12
  catchments in southern New England. The R package can be used to develop N delivery
  maps for a specific HUC-12 of interest, or users can visit the University of Connecticut
  website to explore previously mapped areas. Users can also investigate specific
  flowpaths interactively by clicking on any location within the catchment. A flowpath
  is generated, along with a table describing N removal along each segment."
---

@@

```{r setup, echo=FALSE, warning = FALSE}
library(FedData)
library(nsink)
library(raster)
library(ggplot2)
library(terra)
library(tidyterra)
```

@@

## Development of the R package `nsink`
The package `nsink` implements the approach detailed in Kellogg et al. [-@kellogg2010geospatial] to estimate relative nitrogen (N) removal along a flowpath. The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder. The initial version required time-consuming data manipulation by hand due to limitations of earlier NHDPlus datasets as well as the limitations of working in a GIS environment requiring a user license. With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink was automated and applied to other HUC-12 catchments, leading to the development of the `nsink` package. Specifically, `nsink` relies on packages `sf` and `raster`. The `nsink` package is available from https://github.com/usepa/nsink and is fully described in [@hollister2022nsink].

@@

These static maps can be quickly plotted with the convenience function `nsink_plot()` using the `map` argument to identify map type. For example, `nsink_plot(niantic_static_maps, map = "removal")` will create a plot of Nitrogen Removal Efficiency (Figure 1).  The other `map` argument options are "transport" and "delivery" (Figures 2-3). These plots are designed for exploration and not necessarily for high quality maps.  For higher quality maps, the source data can be plotted in software of a users choice (e.g. QGIS, `ggplot2`).

@@

```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (Table 3)."}

@@

```{r table3-build, eval = TRUE, include = FALSE}
table3 <- data.frame("Land Cover" = c("Developed", "Agriculture", "Forest/Vegetated",
                                      "Wetlands", "Other"),
                     "West (%)" = c("23","7","64","6","0"),
                     "East (%)" = c("20","0","61","18","1"),
                     check.names = FALSE)
```
```{r table3, include = TRUE, echo = FALSE}
knitr::kable(table3, caption = "Table 3: Summary of Land Cover in Highlighted Areas of Interest")
```

@@

Many people have contributed in various ways to the development of the N-Sink 
concept. In particular, we would like to thank Peter August and Chris Damon of 
the University of Rhode Island’s Department of Natural Resources Science.  
Additionally, we are grateful to Stephen Shivers, Anne Kuhn, Andrea Bartolotti, 
Suzanne Ayvazian, and Tim Gleason for constructive reviews of this paper. The 
views expressed in this article are those of the authors and do not necessarily 
represent the views or policies of the U.S. Environmental Protection Agency. Any 
mention of trade names, products, or services does not imply an endorsement by 
the U.S. Government or the U.S. Environmental Protection Agency. The EPA does 
not endorse any commercial products, services, or enterprises. This contribution 
is identified by the tracking number ORD-054342 of the Atlantic Coastal 
Environmental Sciences Division, Office of Research and Development, Center for 
Environmental Measurement and Modeling, US Environmental Protection Agency.

>>>>

---
title: "Assessing landscape N removal in coastal New England catchments using the
  N-Sink approach with a new R package, nsink"
author:
- "Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\\*^ ^*2*^, Arnold, C. L.^*3*^, Gold, A.
  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Lei-Parent,
  Qian*^*3*^, Forshay, K. J.^*4*^"
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  pdf_document: default
  word_document:
    reference_docx: manuscript_template.docx
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: "Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water
  quality. Land management decisions made within estuarine watersheds have a direct
  impact on downstream N delivery. Natural features frequently found within the headwaters
  of watersheds can act as landscape sinks for N, such as wetlands, streams and ponds
  that transform dissolved N into gaseous N, effectively removing it from the aquatic
  system. Identifying and evaluating these landscape sinks and their spatial relationship
  to N sources can help managers understand the effects of alternative decisions on
  downstream resources. N-Sink is an approach that uses widely available GIS data
  to identify landscape sinks within HUC-12 (or larger) catchments, estimate their
  N removal potential and map the effect of those sinks on N movement through the
  catchment. A series of static maps are produced to visualize N removal efficiency,
  transport and delivery, the latter in the form of an index. To facilitate data acquisition,
  processing and visualization, an R package `nsink` was developed and applied to
  produce static maps for all HUC-12 catchments in southern New England. The R package
  can be used to develop N delivery maps for a specific HUC-12 of interest, or users
  can visit the University of Connecticut website to explore previously mapped areas.
  Users can also investigate specific flowpaths interactively by clicking on any location
  within the catchment. A flowpath is generated, along with a table describing N removal
  along each segment."
---

@@

```{r setup, echo=FALSE, eval = TRUE, warning = FALSE}
library(FedData)
library(nsink)
library(raster)
library(ggplot2)
library(terra)
library(tidyterra)
```

@@

## Development of the R package `nsink`
The package `nsink` implements the approach detailed in Kellogg et al. [-@kellogg2010geospatial] to estimate relative nitrogen (N) removal along a flowpath. The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder. The initial version required time-consuming data manipulation by hand due to limitations of earlier NHDPlus datasets as well as the limitations of working in a GIS environment requiring a user license. With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink was automated and applied to other HUC-12 catchments, leading to the development of the `nsink` package. Specifically, `nsink` relies on packages `sf` and `raster`. The `raster` package in R will soon be superseded by the `terra` package, which will be integrated in future `nsink` package versions. The `nsink` package is available from https://github.com/usepa/nsink and is fully described in [@hollister2022nsink].

@@

These static maps can be quickly plotted with the convenience function `nsink_plot()` using the `map` argument to identify map type. For example, `nsink_plot(niantic_static_maps, map = "removal")` will create a plot of Nitrogen Removal Efficiency (Figure 1).  The other `map` argument options are "transport" and "delivery" (Figures 2-3). These plots are designed for exploration and not necessarily for high quality maps.  For higher quality maps, the source data can be plotted in software of a user's choice (e.g. QGIS, `ggplot2`).

@@

```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (Table 4)."}

@@

```{r table4-build, eval = TRUE, include = FALSE}
table3 <- data.frame("Land Cover" = c("Developed", "Agriculture", "Forest/Vegetated",
                                      "Wetlands", "Other"),
                     "West (%)" = c("23","7","64","6","0"),
                     "East (%)" = c("20","0","61","18","1"),
                     check.names = FALSE)
```
```{r table4, include = TRUE, echo = FALSE}
knitr::kable(table3, caption = "Table 3: Summary of Land Cover in Highlighted Areas of Interest")
```

@@

Many people have contributed in various ways to the development of the N-Sink 
concept. In particular, we would like to thank Peter August and Chris Damon of 
the University of Rhode Island’s Department of Natural Resources Science.  
Additionally, we are grateful to Stephen Shivers, Anne Kuhn, Andrea Bartolotti, 
Suzanne Ayvazian, and Tim Gleason for constructive reviews of this paper.

The views expressed in this article are those of the authors and do not necessarily 
represent the views or policies of the U.S. Environmental Protection Agency. Any 
mention of trade names, products, or services does not imply an endorsement by 
the U.S. Government or the U.S. Environmental Protection Agency. The EPA does 
not endorse any commercial products, services, or enterprises. This research has 
been supported, in part, by a cooperative agreement with the U.S. Environmental
Protection Agency's Office of Research and Development, EPA Assistant Agreement
Number CR83680001, "N-Sink tool and outreach to two watersheds in New England."
This contribution is identified by the tracking number ORD-054342 of the 
Atlantic Coastal Environmental Sciences Division, Office of Research and Development, 
Center for Environmental Measurement and Modeling, US Environmental Protection Agency.
